Maximum Entropy Spectral Analysis: a case study

### 2021-02-25 paper/mesa.tex
Validation part completed

We will apply Burg method with each loss function on a number of simulated time series, obtaining a variety of estimate for PSD.
The disagreement between the PSD $S_i(f)$ estimated from the i-th simulated time series and the target PSD $S(f)$ is measured with the relative error $r_i$, averaged over all frequencies:
\begin{equation}
r_i = \frac{1}{f_{Ny}}\int_0^{f_{Ny}} \d{f} \frac{|S_i(f) - S(f)|}{S(f)}
\end{equation}
where $f_{Ny}$ is the Nyquist frequency of the time series. Of course in practice, the integral will be computed by averaging over a discrete grid of frequencies.

@@

\subsection{Choosing a loss function - LIGO Spectrum} \label{sec:LIGO_validation}
\begin{figure}
    \centering
    \includegraphics[width = \linewidth]{Images/LIGOsimulate/LigoPSDoriginal.pdf}
    \caption{Original power spectral density from Ligo}
    \label{fig:LigoPSDOriginal}
\end{figure}
In this subsection we concentrate again on the reconstruction of a specific, known, power spectrum.
We analyze the properties of reconstruction of this second spectrum whose shape is the Advanced LIGO design sensitivity theoretical spectral curve (Figure \ref{fig:LigoPSDOriginal}). 
For this analysis we will study properties only for arrays of 40960 points. We consider a sampling rate of $2048 samples / second$ that means we can reconstruct a Nyquist frequency of $1024 Hz$. We chose a longer dataset than before because algorithm works with a 1 dimensional interpolation of spectrum, and with shorter datasets the interpolated spectrum is not consistent with the original spectrum and does not capture the height of the peaks. Shorter choices would result in noise extracted from a different power spectrum than the one in figure \ref{fig:LigoPSDOriginal}. \\ 
We will study statistical properties from samples obtained with 500 simulations and compare the obtained results for each optimizer. 
We start considering statistical properties of mean spectrum obtained averaging over all the simulations and conclude with an analysis for the statistics of the reproduction of single spectrum. 
\subsubsection{Properties for the average spectrum}
Here we report mean properties of the reproductions of the spectrum. The spectrum we want to reconstruct is reported in Figure \ref{fig:ligospectrum},
\begin{figure}
    \centering

        \includegraphics[width = \linewidth]{Images/LIGOsimulate/PSD.pdf}
        \caption{A priori spectrum from Ligo database}
        \label{fig:ligospectrum}
shows the original power spectral density and the ensemble average of every method. For the mean, the relative errors are: 
    \bar r_{\bar S, FPE} = 0.0426\%, \qquad \bar r_{\bar S, OBD} = 0.129\%, \qquad \bar r_{\bar S, CAT} = 0.0211\%
Again, if one only check overall accuracy for every method, CAT is outperforming. In fact, even in presence of very sharp peaks, CAT reconstruction seems to be almost perfectly coincident with each of them. To check this behaviour we take a look at the two peaks separately. For the first one a log plot of the reconstructed peaks and their relative errors are shown in Figure \ref{fig:ligo1peak} 
\begin{figure}[t]
        \includegraphics[width = \linewidth]{Images/LIGOsimulate/1stPeakComparison.pdf}
        \caption{1st peak of the spectrum and it reconstruction with every optimizer (left) and associated percentage error (right)}
        \label{fig:ligo1peak}
and the results reinforce our previous statement. The behaviour of the error is very similar for CAT and FPE, showing similar shapes but very different sizes, while CAT is giving a different and way more precise result. Both FPE and OBD are not able to capture the width of the peak and also show larger errors, while CAT has a good behaviour in reconstructing width's also. 
The reconstruction for very sharp peaks considering the average over all the spectra is found to be strongly inaccurate for both FPE and OBD, while a good accuracy is reached with CAT method. \\ 
Considering the estimate for the position of the peak, whose true value is found to be $16.85 \quad a.u. $, within numerical accuracy both FPE and CAT find its exact position, while OBD shift the peak left at $16.7 a.u.$, with a $0.9\%$ inaccuracy.\\ 
Second peak results are analogous, plotting the peak and the errors value around it
\begin{figure}[t]
    \centering
        \includegraphics[width = \linewidth]{Images/LIGOsimulate/2ndPeakComparison.pdf}
        \caption{2nd peak of the spectrum, Relative error around 2nd peak of the spectrum}
        \label{fig:ligo2peak}
we find again that FPE and OBD are performing similarly, the magnitude of the error is maximum for OBD and minimum for CAT. \\ 
In this case even CAT shows large errors in some points, but they are reasonable in a large part of it, still, it remains the best reconstruction.  For this second peak, all the methods find maximum position to be at $438.05 \quad a.u.$, which is the true position of the maximum.
Looking at Figure \ref{fig:ligo1peak} at \ref{fig:ligo2peak}, it appears to be clear that the least of the accuracy is not on the peak, but immediately around, so that one of the biggest deal is the reconstruction of peak's width, that only CAT is able to capture with high definition. \\ 
Even if FPE shows larger error in the peaks than CAT, when the average over all the frequencies is considered, their percentage errors are of the same order of magnitude. This means that outside the peaks FPE is more accurate than CAT, i.e. CAT is still giving noisy results.\\ 
As for the normal distributed noise, mean spectrum is not sufficient to declare what is the best optimizer to be used, since most of experiment are purely observational and mean spectrum over different realizations of the same stochastic process cannot be computed. So, it is important to also understand what is their behaviour for single realizations of the process.
\subsubsection{Statistical properties of single realizations} 
The errors for the single realizations behave very differently from the errors obtained when we considered the mean spectrum. In computing the mean of the overall errors for each single spectrum, the results we obtain are the following:
    \bar{r}_{FPE} = 0.13 \pm 0.01; \qquad \bar{r}_{OBD} = 0.17 \pm .02; \qquad 
    \bar{r}_{CAT} = 0.4 \pm 0.1
where the error is just the standard deviation of the samples.
In this case we find a different result: FPE shows the best results and order's estimate is associated with least errors. Also, having found a smaller standard deviation of the datas, we can state that FPE and OBD are more stable optimizers then CAT. In fact, their properties as a function of filter's length are almost identical to those obtained in the study of normal spectrum. If we again plot the value of the errors as a function of the estimate of the order (Figure \ref{fig:LigoOrderError})
    \centering
    \includegraphics[width = \linewidth]{Images/optimisers_comparison/ligo/error_length_contour.pdf}
    \caption{Plot of relative error as a function of the estimate for filter's length}
    \label{fig:LigoOrderError}
it is immediate to see that, again, FPE results lies in the are of filter's length associated with a minimum of relative errors, OBD lies in a clustered window with a higher error and CAT is not convergent, but most of the results are outside the minimum error area. So, we arrive at the same conclusion we had when we considered the normal specturm. OBD has no general properties that make it a preferable optimizer then FPE or CAT while, what to choose beneath the last two is based on the problem we are dealing with: if we have one single realization for the process, we are mostly sure that FPE would catch the area associated with the best resolution possible, while CAT would result in a spurious result with, in general, a not so good resolution. But, in case we have several realization for the same process, this last property of CAT, together with its instability, result to be his luck and, taking the average errors tend to compensate noise in the result, originating a very well defined spectrum that show same resolution almost everywhere, no matter how small the values are (as for normal PSD case) or how sharp peaks can be, the result has got generally good accuracy, also if it is clear that we have to pay this fact with very large error bars. So a choice between these methods is a matter of convenience and interest for each specific problem. \\

>>>>

We will apply Burg method with each loss function on a number of simulated time series, obtaining a variety of estimate for PSD.
The disagreement between the PSD $S_i(f)$ estimated from the i-th simulated time series and the target PSD $S(f)$ is measured with the relative error $r_i$, averaged over all frequencies:
\begin{equation}\label{eq:freq_erorr}
r_i = \frac{1}{f_{Ny}}\int_0^{f_{Ny}} \d{f} \frac{|S_i(f) - S(f)|}{S(f)}
\end{equation}
where $f_{Ny}$ is the Nyquist frequency of the time series. Of course in practice, the integral will be computed by averaging over a discrete grid of frequencies.

@@

\subsection{Choosing a loss function - LIGO Spectrum} \label{sec:LIGO_validation}
\begin{figure}
    \centering
    \includegraphics[width = \linewidth]{Images/LIGOsimulate/LigoPSDoriginal.pdf}
    \caption{Original power spectral density from Ligo}
    \label{fig:LigoPSDOriginal}
\end{figure}
In this subsection we concentrate again on the reconstruction of a specific, known, power spectrum: that is the Advanced LIGO design sensitivity theoretical spectral curve see fig.~\ref{fig:LigoPSDOriginal}.
For this analysis we will study properties on $500$ time series arrays of $40960$ points each. We set a sampling rate of $\SI{2048}{Hz}$ (this is $\SI{20}{s} of data$).
We chose a longer dataset to capture the complicated structure of the PSD. A shorter time batch would not have allowed us to reconstruct faithfully all the peaks of the empirical PSD.

\begin{figure}
    \centering
     \includegraphics[width = \linewidth]{Images/LIGOsimulate/PSD.pdf}
      \caption{A priori spectrum from Ligo database}
       \label{fig:ligospectrum}
\end{figure}
In figure \ref{fig:ligospectrum}, we report the original spectrum and the ensemble averaged spectrum obtained with the three methods (no confidence level).
In figure \ref{fig:ligo1peak} and \ref{fig:ligo2peak}, we plot a details od fig.~\ref{fig:ligospectrum} around the two peaks of the spectrum
In figure~\ref{fig:LigoOrderError}, we report, as above, a scatter plot with the frequency averaged error (eq.~\ref{eq:freq_erorr}) and the length of the autoregressive model obtained with each loss function.

The conclusions are very similar to those above. When averaged on multiple realization, CAT loss function is able to capture the spectrum very precisely.
In fact, even in presence of very sharp peaks, CAT reconstruction seems to be almost perfectly coincident with each of them (figs.\ref{fig:ligo1peak} and \ref{fig:ligo2peak}).
However, the computation on a single noise realization yields larger error, as manifest in fig.~\ref{fig:LigoOrderError}.
This supports the conclusion that CAT is not a stable optimizer: different computation of the same spectrum can output very different results.
On the other hand, FPE seems more robust and provides a smaller variance (and frequency averaged error), despite having a worse performance on the average.
We find again that FPE and OBD have similar performance: indeed, OBD shows qualitatively the same behaviour of FPE, although with slightly worse performance in the error.

We can then draw some general conclusions:
\begin{itemize}
	\item There are no reasons to prefer OBD over CAT or FPE.
	\item If we have one single realization for the process, we recommend the use of FPE, that would get the best resolution possible. In this situation, CAT would provide spurious and unreliable results, with large error.
	\item But, in case we have several realization for the same process, this last property of CAT, together with its instability, result to be his luck: averaging the spectrum over different time series tend to compensate noise in the results, originating a very well precise spectral estimation. The same is not true for FPE.
\end{itemize}

The choice of the loss function between CAT and FPE seems problem-dependent.

\begin{figure}[t]
        \includegraphics[width = \linewidth]{Images/LIGOsimulate/1stPeakComparison.pdf}
        \caption{1st peak of the spectrum and it reconstruction with every optimizer (left) and associated percentage error (right)}
        \label{fig:ligo1peak}
\end{figure}
\begin{figure}[t]
    \centering
        \includegraphics[width = \linewidth]{Images/LIGOsimulate/2ndPeakComparison.pdf}
        \caption{2nd peak of the spectrum, Relative error around 2nd peak of the spectrum}
        \label{fig:ligo2peak}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width = \linewidth]{Images/optimisers_comparison/ligo/error_length_contour.pdf}
    \caption{Plot of relative error as a function of the estimate for filter's length}
    \label{fig:LigoOrderError}
\end{figure}
